\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{authblk}

\title{\textbf{Adapting Vision Foundation Models for Semantic Correspondence: A Study on PEFT and Adaptive Spatial Refinement}}

\author[1]{Johnprice Osagie}
\author[1]{Mario Lapadula}
\author[1]{Giorgia Pugliese}
\author[1]{Riccardo Bellanca}
\affil[1]{Politecnico di Torino, Corso di Laurea in Ingegneria Informatica}

\date{Academic Year 2023/2024}

\begin{document}

\maketitle

\begin{abstract}
Establishing dense semantic correspondences remains a fundamental challenge in computer vision. While large-scale Foundation Models like DINOv2 provide robust feature representations, their zero-shot application to part-level matching often suffers from spatial imprecision. In this work, we propose a two-pronged adaptation strategy. First, we explore Parameter-Efficient Fine-Tuning (PEFT) using LoRA and BitFit to specialize the backbone for coordinate regression. Second, we introduce an entropy-calibrated Adaptive Window Soft-Argmax stage to refine spatial localization. Our final pipeline (LoRA + AW) achieves 80.53\% PCK@0.10 on the SPair-71k dataset, outperforming the zero-shot baseline by a significant margin (+36\%) while training only 1.03\% of the model parameters.
\end{abstract}

\section{Introduction}
Semantic correspondence concerns the task of establishing point-to-point matches between two images of different instances belonging to the same semantic category. Unlike classic geometric matching, semantic correspondence must handle high intra-class variations, occlusions, and non-rigid deformations. 

Recent advances in Self-Supervised Learning (SSL), particularly DINOv2, have demonstrated that Vision Transformers (ViTs) can learn object-centric representations without human supervision. However, extracting sub-pixel accurate correspondences from these models requires careful calibration. In this report, we analyze the impact of efficient adaptation techniques and spatial filtering mechanisms to maximize matching precision.

\section{Methodology}
\subsection{Backbone: DINOv2}
We employ DINOv2 (ViT-B/14) as our primary feature extractor. Let $I_s$ and $I_t$ be the source and target images. We extract features $F_s, F_t \in \mathbb{R}^{H \times W \times D}$ from a specific layer $L$. Our empirical analysis (Sec. \ref{sec:layerwise}) identifies the penultimate layer ($L=11$) as the most semantically rich.

\subsection{Low-Rank Adaptation (LoRA)}
To adapt the backbone efficiently, we freeze the original weights $W_0$ and inject trainable low-rank matrices. For a linear layer $y = W_0 x$, the modified forward pass is:
\begin{equation}
    y = W_0 x + \frac{\gamma}{r} (A B) x
\end{equation}
where $A \in \mathbb{R}^{d \times r}$ and $B \in \mathbb{R}^{r \times k}$ are low-rank matrices ($r=16$), and $\gamma$ is a scaling factor. This allows the model to learn task-specific semantic shifts without full re-training.

\subsection{BitFit Analysis}
As a baseline for efficiency, we also evaluate BitFit, which limits training to only the bias terms of the network:
\begin{equation}
    y = W_0 x + (b_0 + \Delta b)
\end{equation}
Surprisingly, BitFit achieves competitive results (Table \ref{tab:pck}), proving that many semantic shifts in foundation models can be captured by simple bias translations.

\subsection{Spatial Refinement: Adaptive Windowing}
The standard correlation map $C = \text{cosine\_sim}(F_s, F_t)$ often contains secondary peaks due to background clutter. We propose an \textbf{Adaptive Window (AW)} approach. Before the Soft-Argmax, we filter the scores:
\begin{equation}
    C'_{i,j} = 
    \begin{cases} 
    C_{i,j}, & \text{if } C_{i,j} > \tau \\
    -\infty, & \text{otherwise}
    \end{cases}
\end{equation}
where $\tau$ is a quantile-based threshold. This limits the localized soft-weighted average to the primary semantic neighborhood.

\section{Experimental Setup}
\begin{itemize}
    \item \textbf{Datasets}: SPair-71k (Main), PF-Pascal (OOD testing).
    \item \textbf{Hardware}: Google Colab (Tesla T4/L4 GPUs).
    \item \textbf{Hyperparameters}: Learning rate $1 \times 10^{-4}$, 5 epochs, AdamW optimizer, batch size 16.
\end{itemize}

\section{Results}
\subsection{Zero-Shot Baselines}
Comparing DINOv2 with DINOv3 and SAM (Segment Anything Model) reveals a stark difference in semantic awareness. SAM, while great for segmentation, lacks the discriminative features needed for correspondence (PCK < 20\%).
% PLACEHOLDER: fig:baselines
% [IMAGE: 01_Baselines_SPair.png]

\subsection{Quantitative Results}
As shown in Table \ref{tab:pck}, the LoRA + AW combination yields the highest precision, particularly in the strict PCK@0.05 regime.

\begin{table}[h]
\centering
\caption{PCK evaluation on SPair-71k test set.}
\label{tab:pck}
\begin{tabular}{@{}lcc@{}}
\toprule
Architecture & PCK@0.10 & PCK@0.05 \\ \midrule
Baseline DINOv2 & 43.88\% & 24.28\% \\
BitFit (Smooth) & 75.40\% & 50.81\% \\
BitFit + AW & 77.83\% & 55.48\% \\
LoRA (Standard)  & 78.13\% & 55.40\% \\
\textbf{LoRA + AW} & \textbf{80.53\%} & \textbf{61.36\%} \\ \bottomrule
\end{tabular}
\end{table}

\subsection{Temperature and Regularization}
Temperature scaling ($T$) is used to sharpen the Soft-Argmax distribution. $T \to 0$ approaches Hard-Argmax, while $T=0.05$ (our choice) provides a soft-spatial regularizer that enables sub-pixel accuracy while maintaining differentiability.
% PLACEHOLDER: fig:temp
% [IMAGE: 07_Temperature_Calibration.png]

\subsection{Layer-wise Analysis} \label{sec:layerwise}
Lower layers (L4, L8) capture textures and parts but lack the global category knowledge of L11.
% PLACEHOLDER: fig:layerwise
% [IMAGE: 09_LayerWise_Analysis.png]

\subsection{Robustness and Generalization}
The model maintains a +20\% lead over frozen baselines on PF-Pascal, proving that the PEFT adaptation is not dataset-overfitted but semantically general. Synthetic rotation tests show that while ViT backbones degrade under 90°/180° rotations, the adapted layers provide a higher stability baseline.
% PLACEHOLDER: fig:robustness
% [IMAGE: 06_Robustness_Rotation.png]

\section{Qualitative Analysis and Demo}
We developed an interactive Gradio dashboard to visualize results. Our qualitative analysis confirms that the Adaptive Window successfully suppresses "ghosting" artifacts in images with complex backgrounds.
% PLACEHOLDER: fig:qualitative_collage
% [IMAGE: Gradio_Captures/collage.png]

\section{Future Work}
Future research should focus on \textbf{V-JEPA} (Vision Joint-Embedding Predictive Architecture). Unlike DINOv2, JEPA-style training aims at learning world dynamics and causal relationships, which could drastically improve semantic correspondence in dynamic videos or under extreme occlusion.

\section{Conclusion}
By combining low-rank adapters with adaptive spatial filtering, we demonstrate that foundation models can be turned into highly accurate semantic matchers with negligible computational cost.

\end{document}
